---
title: "R_For_Data_Science"
author: "Maximilian Greil"
date: "11 4 2021"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    theme: journal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r}
suppressMessages(library(tidyverse))
suppressMessages(library(downloader))
suppressMessages(library(plotly))
suppressMessages(library(car))
suppressMessages(library(dunn.test))
suppressMessages(library(broom))
```

## 1) Read in the <code>gapminder_clean.csv</code> data as a <code>tibble</code> using <code>read_csv</code>

Read in gapminder_clean.csv as a tibble using read_csv

```{r}
dat <- read.csv('gapminder_clean.csv') %>% as_tibble
dim(dat)
```

The gapminder data consists of a matrix with 2607 rows and 20 columns.

## 2) Filter the data to include only rows where Year is <code>1962</code> and then make a scatter plot comparing <code>'CO2 emissions (metric tons per capita)'</code> and <code>gdpPercap</code> for the filtered data

```{r}
dat %>% 
  filter(Year == 1962) %>%
  # remove rows which contain NAs
  filter_at(vars(CO2.emissions..metric.tons.per.capita., gdpPercap), all_vars(!is.na(.))) %>%
  ggplot(aes(x = gdpPercap, y = CO2.emissions..metric.tons.per.capita.)) +
  geom_point() +
  ylab("CO2 emissions (metric tons per capita)")

# Because of one very large outlier the analysis repreated with log transformed data

dat %>% 
  filter(Year == 1962) %>%
  # remove rows which contain NAs
  filter_at(vars(CO2.emissions..metric.tons.per.capita., gdpPercap), all_vars(!is.na(.))) %>%
  ggplot(aes(x = log(gdpPercap), y = log(CO2.emissions..metric.tons.per.capita.))) +
  geom_point() +
  ylab("log(CO2 emissions (metric tons per capita))")
```

Log transformed data can be used to make data as “normal” as possible so that the statistical analysis results become more valid.

## 3) On the filtered data, calculate the pearson correlation of <code>'CO2 emissions (metric tons per capita)'</code> and <code>gdpPercap</code>. What is the Pearson R value and associated p value?

```{r}
dat %>% 
  filter(Year == 1962) %>%
  # remove rows which contain NAs
  filter_at(vars(CO2.emissions..metric.tons.per.capita., gdpPercap), all_vars(!is.na(.))) %>%
  pivot_longer( gdpPercap, names_to = "x_var", values_to = "x_val") %>%
  pivot_longer( CO2.emissions..metric.tons.per.capita., names_to = "y_var", values_to = "y_val") %>%
  group_by(x_var, y_var) %>%
   summarise(cor_coef = cor.test(x_val, y_val)$estimate,
            p_val = cor.test(x_val, y_val)$p.value)
```

The Pearson R value of 'CO2 emissions (metric tons per capita)' and gdpPercap in Year 1962 is 0.9260817 and the associated p value is 1.128679e-46.

```{r}
# Because of one very large outlier the analysis repreated with log transformed data

dat %>% 
  filter(Year == 1962) %>%
  # remove rows which contain NAs
  filter_at(vars(CO2.emissions..metric.tons.per.capita., gdpPercap), all_vars(!is.na(.))) %>%
  pivot_longer( gdpPercap, names_to = "x_var", values_to = "x_val") %>%
  pivot_longer( CO2.emissions..metric.tons.per.capita., names_to = "y_var", values_to = "y_val") %>%
  group_by(x_var, y_var) %>%
  # calculate pearson R value and associated p value
  summarise(cor_coef = cor.test(log(x_val), log(y_val))$estimate,
            p_val = cor.test(log(x_val), log(y_val))$p.value)
```

The Pearson R value of log transformed 'CO2 emissions (metric tons per capita)' and gdpPercap in Year 1962 is 0.8602081 and the associated p value is 8.903567e-33.\

## 4) On the unfiltered data, answer "In what year is the correlation between <code>'CO2 emissions (metric tons per capita)'</code> and <code>gdpPercap</code> the strongest?" Filter the dataset to that year for the next step...

```{r}
year.cor.max <- dat %>%
  # remove rows which contain NAs
  filter_at(vars(CO2.emissions..metric.tons.per.capita., gdpPercap), all_vars(!is.na(.))) %>%
  # split origial tibble in lists depending on the year
  split(.$Year) %>%
  # perform pearson correlation test on each year list, keep only the pearson R value
  map(~cor.test(log(.$CO2.emissions..metric.tons.per.capita.), log(.$gdpPercap))$estimate) %>% 
  # merge list of lists to one final list
  unlist

year.cor.max

year.cor.max[which.max(year.cor.max)]
```

In year 2002 is the correlation between 'CO2 emissions (metric tons per capita)' and gdpPercap the strongest.

## 5) Using <code>plotly</code>, create an interactive scatter plot comparing <code>'CO2 emissions (metric tons per capita)'</code> and <code>gdpPercap</code>, where the point size is determined by <code>pop</code> (population) and the color is determined by the <code>continent</code>. You can easily convert any <code>ggplot</code> plot to a <code>plotly</code> plot using the <code>ggplotly()</code> command.

```{r}
viz <- dat %>% 
    filter(Year == 2002) %>%
    # remove rows which contain NAs
    filter_at(vars(gdpPercap, CO2.emissions..metric.tons.per.capita.), all_vars(!is.na(.))) %>%
    ggplot(aes(x = log(gdpPercap), y = log(CO2.emissions..metric.tons.per.capita.), text=paste0('</br>log(gdpPercap): ', gdpPercap, '</br>log(CO2 emissions (metric tons per capita)): ', CO2.emissions..metric.tons.per.capita., '</br>population: ', pop, '</br>continent: ', continent, '</br>Country Name: ', Country.Name))) +
    geom_point(aes(size = pop, color = continent)) +
    ylab("log(CO2 emissions (metric tons per capita))")
  
ggplotly(viz, tooltip = c("text"))
```

## 6) What is the relationship between <code>continent</code> and <code>'Energy use (kg of oil equivalent per capita)'</code>?

First we must choose a suitable statistical test. We have continuous data and want to check for differences in mean for more than two groups.\ 

Therefore, we have to check if the parametric assumptions are satisfied. For ANOVA, we have four parametric assumptions that must be met:

- Samples must be independent
- Population variances must be equal (Levene’s test)
- The population must be close to a normal distribution (Shapiro-Wilk normality test)
- Groups must have equal sample sizes

### 6.1) Samples must be independent

```{r}
dat %>%
  filter(continent != "") %>%
  # remove rows which contain NAs
  filter_at(vars(Energy.use..kg.of.oil.equivalent.per.capita.), all_vars(!is.na(.))) %>%
  ggplot(aes(x=continent, y=Energy.use..kg.of.oil.equivalent.per.capita., fill = continent)) +
  geom_boxplot() +
  ylab("Energy use (kg of oil equivalent per capita)")
```

First, exploratory data analysis (EDA) was performed on the data. We can say that the samples are independent.

### 6.2) Population variances must be equal (Levene’s test)

```{r}
dat %>%
  filter(continent != "") %>%
  # remove rows which contain NAs
  filter_at(vars(Energy.use..kg.of.oil.equivalent.per.capita.), all_vars(!is.na(.))) %>%
  # perform Levene’s test
  do(tidy(leveneTest(.$Energy.use..kg.of.oil.equivalent.per.capita.~.$continent, data = .)))
```

We reject the H0 hypothesis that the variances are equal based on the p value of our Levene’s test. Therefore, we can not use an ANOVA test and must use a Kruskal-Wallis test instead.

### 6.3) Kruskal-Wallis test

```{r}
dat %>%
  filter(continent != "") %>%
  # remove rows which contain NAs
  filter_at(vars(Energy.use..kg.of.oil.equivalent.per.capita.), all_vars(!is.na(.))) %>%
  # perform Kruskal-Wallis test
  do(tidy(kruskal.test(.$Energy.use..kg.of.oil.equivalent.per.capita.~.$continent, data = .)))
```  

The resulting p value of our Kruskal-Wallis test shows that there is a significant difference in the mean for our data. Because of our significant result, we additionally perform a dunn's test.

```{r}
dat6 <- dat %>%
  filter(continent != "") %>%
  # remove rows which contain NAs
  filter_at(vars(Energy.use..kg.of.oil.equivalent.per.capita.), all_vars(!is.na(.)))

# perform dunn's test
dunn.test(dat6$Energy.use..kg.of.oil.equivalent.per.capita., g=dat6$continent)
```  

## 7) Is there a significant difference between Europe and Asia with respect to <code>'Imports of goods and services (% of GDP)'</code> in the years after 1990?

Again, we must choose a suitable test first. We have continuous data and want to check for differences in mean for two groups.\ 

Therefore, we have to check if the parametric assumptions are satisfied. For Student's unpaired t-test, we have four parametric assumptions that must be met:

- The observations are sampled independently
- The dependent variable is measured on an incremental level, such as ratios or intervals.
- The independent variables must consist of two related groups or matched pairs.
- The dependent variable is normally distributed (Shapiro-Wilk normality test)

### 7.1) The observations are sampled independently

```{r}
dat %>%
  filter(Year > 1990 & (continent == "Europe" | continent == "Asia")) %>%
  # remove rows which contain NAs
  filter_at(vars(Imports.of.goods.and.services....of.GDP.), all_vars(!is.na(.))) %>%
  ggplot(aes(x=continent, y=Imports.of.goods.and.services....of.GDP., fill = continent)) +
  geom_boxplot() +
  ylab("Imports of goods and services (% of GDP)")
```

First, exploratory data analysis (EDA) was performed on the data. We can say that the samples are independent. We can also say that the dependent variable is measured on the incremental level year and that independent variables consist of matched pairs.

### 7.2) The dependent variable is normally distributed (Shapiro-Wilk normality test)

```{r}
dat %>%
  filter(Year > 1990 & (continent == "Europe" | continent == "Asia")) %>%
  # remove rows which contain NAs
  filter_at(vars(Imports.of.goods.and.services....of.GDP.), all_vars(!is.na(.))) %>%
  # perform Shapiro-Wilk normality test
  do(tidy(shapiro.test(.$Imports.of.goods.and.services....of.GDP.)))
```

We reject the H0 hypothesis that the data is normally distributed based on the p values of our Shapiro-Wilk normality test. Therefore, we can not use a Student's unpaired t-test and must use a Wilcoxon Rank sums test instead.

### 7.3) Wilcoxon Rank sums test

```{r}
dat %>%
  filter(Year > 1990 & (continent == "Europe" | continent == "Asia")) %>%
  # remove rows which contain NAs
  filter_at(vars(Imports.of.goods.and.services....of.GDP.), all_vars(!is.na(.))) %>%
  # perform Wilcoxon Rank sums test
  do(tidy(wilcox.test(.$Imports.of.goods.and.services....of.GDP.~.$continent)))
```

We reject the H0 hypothesis that there is a significant difference between Europe and Asia with respect to 'Imports of goods and services (% of GDP)' in the years after 1990 based on the p values of our Wilcoxon Rank sums test.

## 8) What is the country (or countries) that has the highest <code>'Population density (people per sq. km of land area)'</code> across all years? (i.e., which country has the highest average ranking in this category across each time point in the dataset?)

First, exploratory data analysis (EDA) was performed on the data.

```{r}
viz <- dat %>%
  # remove rows which contain NAs
  filter_at(vars(Year, Country.Name, Population.density..people.per.sq..km.of.land.area.), all_vars(!is.na(.))) %>%
  ggplot(aes(x=Year, y=Population.density..people.per.sq..km.of.land.area., group = Country.Name)) + 
  geom_line() + 
  ylab("Population density (people per sq. km of land area)")

ggplotly(viz)
```

```{r}
highest.number <- dat %>%
  # remove rows which contain NAs
  filter_at(vars(Year, Country.Name, Population.density..people.per.sq..km.of.land.area.), all_vars(!is.na(.))) %>%
  # split origial tibble in lists depending on the year
  split(.$Year) %>%
  # get maximum of variable in every year
  map(~max(.$Population.density..people.per.sq..km.of.land.area.)) %>%
  # merge list of lists into one list
  unlist

# map highest scoring variable from every year to respective country
highest.ranked.countries <- data.frame(lapply(highest.number, function(x) dat$Country.Name[which(dat$Population.density..people.per.sq..km.of.land.area. == x)]), stringsAsFactors=FALSE)

# count occurences of highest scoring countries
apply(highest.ranked.countries, MARGIN=1, table)
```

The countries that have the highest 'Population density (people per sq. km of land area)' across all years are Macao SAR and Monaco.

## 9) What country (or countries) has shown the greatest increase in <code>'Life expectancy at birth, total (years)'</code> since 1962?

First, exploratory data analysis (EDA) was performed on the data.

```{r}
viz <- dat %>%
  # remove rows which contain NAs
  filter_at(vars(Year, Country.Name, Life.expectancy.at.birth..total..years.), all_vars(!is.na(.))) %>%
  ggplot(aes(x=Year, y=Life.expectancy.at.birth..total..years., group = Country.Name)) + 
  geom_line() + 
  ylab("Life expectancy at birth, total (years)")

ggplotly(viz)
```

To get the difference for Life expectancy at birth, total (years) for every country between 1962 and 2007 we subtract the value from 2007 minus the value from 1962.

```{r}
dat %>%
  filter(Year == 1962 | Year == 2007) %>%
  # remove rows which contain NAs
  filter_at(vars(Year, Country.Name, Life.expectancy.at.birth..total..years.), all_vars(!is.na(.))) %>%
  group_by(Country.Name) %>%
  # Calculate difference between 1962 and 2007
  mutate(diff = Life.expectancy.at.birth..total..years. - lag(Life.expectancy.at.birth..total..years., default = 0)) %>% 
  select(Country.Name, diff) %>%
  # keep only rows with calculated difference
  filter(row_number() %% 2 == 0) %>%
  # sort by difference
  arrange(desc(diff))
```

Country Maldives has shown the greatest increase in 'Life expectancy at birth, total (years)' since 1962.